---
title: "Medicare Deaths"
author: "Andrew Bruce"
format: 
  html:
    theme: zephyr
    toc: true
    number-sections: false
date: last-modified
code-fold: true
code-overflow: scroll
---

```{r setup}
#| echo: false
#| eval: true
#| message: false
source(here::here("R", "utils.R"))
source(here::here("R", "deaths.R"))
library(tidyverse)
library(scalesextra)
```


## Medicare Deaths

Information on all Medicare beneficiary deaths, by type of entitlement, health care delivery, demographic characteristics, and geography.

### Deaths by Month

```{r}
#| message: false
# Total Beneficiary Deaths by Month, 2008-2021
mon_tot <- mdcr_enroll_ab_33 |> 
  dplyr::mutate(group = "Total", 
                .after = year)
mon_tot
```


```{r}
# Original Medicare Beneficiary Deaths by Month, 2008-2021
mon_orig <- mdcr_enroll_ab_36 |> 
  dplyr::mutate(group = "Original", 
                .after = year)
mon_orig
```


```{r}
# Medicare Advantage and Other Health Plan Beneficiary Deaths by Month, 2008-2021
mon_ma <- mdcr_enroll_ab_38 |> 
  dplyr::mutate(group = "Medicare Advantage", 
                .after = year)
mon_ma
```


```{r}
# ALL Deaths by Month, 2008-2021
deaths_by_month <- dplyr::full_join(mon_tot, mon_orig) |> 
  dplyr::full_join(mon_ma)
deaths_by_month
```


```{r}
deaths_by_month_pct <- deaths_by_month |> 
  dplyr::select(year, 
                group, 
                dplyr::contains("pct"),
                -dplyr::contains("chg")) |> 
  tidyr::pivot_longer(cols = dplyr::contains("pct"),
                      names_to = "month",
                      values_to = "deaths_pct") |> 
  dplyr::mutate(month = stringr::str_remove(month, "_pct"),
                month = stringr::str_to_title(month),
                month = factor(month, levels = month.name),
                date = clock::as_date(clock::year_month_day(year, as.integer(month), 1)),
                mon = lubridate::month(date, label = TRUE, abbr = TRUE)) |> 
  dplyr::select(date, month, mon, year, group, deaths_pct)
deaths_by_month_pct
```


```{r}
deaths_by_month_n <- deaths_by_month |> 
  dplyr::select(year, 
                group, 
                dplyr::any_of(tolower(month.name))) |> 
  tidyr::pivot_longer(cols = dplyr::any_of(tolower(month.name)),
                      names_to = "month",
                      values_to = "deaths") |> 
  dplyr::mutate(month = stringr::str_to_title(month),
                month = factor(month, levels = month.name),
                date = clock::as_date(clock::year_month_day(year, as.integer(month), 1)),
                mon = lubridate::month(date, label = TRUE, abbr = TRUE)) |> 
  dplyr::select(date, month, mon, year, group, deaths)
deaths_by_month_n
```

```{r}
deaths_by_month_n |>  
  ggplot2::ggplot(aes(x = date)) +
  geom_line(aes(y = deaths, color = group, linetype = group), linewidth = 0.5) +
  scale_fill_manual(name = "", values = c("red","steelblue", "darkgreen")) +
  scale_color_manual(name = "", values = c("red","steelblue", "darkgreen")) +
  scale_linetype_manual(name = "", values = c(1, 1, 1))+
  scalesextra::scale_y_pretty() +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "bottom")
```

<br>

```{r}
deaths_by_month_pct |>  
  dplyr::filter(group == "Total") |> 
  ggplot2::ggplot(aes(x = date)) +
  geom_line(aes(y = deaths_pct 
                #color = group, 
                #linetype = group
                ), linewidth = 0.5) +
  scale_fill_manual(name = "", values = c("red","steelblue", "darkgreen")) +
  scale_color_manual(name = "", values = c("red","steelblue", "darkgreen")) +
  scale_linetype_manual(name = "", values = c(1, 1, 1))+
  scalesextra::scale_y_pct() +
  labs(x = NULL, y = NULL) +
  theme(legend.position = "bottom")
```

<br>

```{r}
deaths_by_month_n |>  
  dplyr::filter(group == "Total", year < 2022) |> 
  ggplot2::ggplot(aes(x = mon)) +
  geom_point(aes(y = deaths, color = year, fill = year)) +
  geom_line(aes(y = deaths, group = year, color = year), linewidth = 0.5) +
  scalesextra::scale_y_pretty2() +
  labs(x = NULL, y = NULL) + 
  theme(legend.position = "bottom")
```

<br>

```{r}
deaths_by_month_pct |>  
  dplyr::filter(group == "Total", year < 2022) |> 
  ggplot2::ggplot(aes(x = mon)) +
  geom_point(aes(y = deaths_pct, color = year, fill = year > 2018, shape = year > 2018), size = 2.5) +
  geom_line(aes(y = deaths_pct, group = year, color = year, linetype = year > 2018), linewidth = 0.5) +
  scalesextra::scale_y_pct() +
  labs(x = NULL, y = NULL) + 
  theme(legend.position = "bottom")
```

<br>

### Deaths by Residence

Total Beneficiary Deaths by Area of Residence, 2013-2021

```{r mdcr_enroll_ab_35}
mdcr_enroll_ab_35
```

<br>

Original Medicare Beneficiary Deaths by Area of Residence, 2013-2021

```{r mdcr_enroll_ab_37}
mdcr_enroll_ab_37
```

<br>

Medicare Advantage and Other Health Plan Beneficiary Deaths by Area of
Residence, 2013-2021

```{r mdcr_enroll_ab_39}
mdcr_enroll_ab_39
```


<br>

### Death Demographics

Total Beneficiary Deaths Demographics, 2013-2021

```{r mdcr_enroll_ab_34}
mdcr_enroll_ab_34
```