---
title: "Medicare Enrollment"
author: "Andrew Burce"
format: 
  html:
    theme: zephyr
    toc: true
    number-sections: false
date: last-modified
code-fold: true
code-overflow: scroll
---

The Medicare Enrollment reports contain trend, demographic, and state tables showing total Medicare enrollment, Original Medicare enrollment, Medicare Advantage and Other Health Plan enrollment, newly-enrolled beneficiaries, deaths, Medicare-Medicaid Enrollment, and Medicare Part D enrollment. These include data on all Medicare enrollees by geography, type of coverage, type of entitlement, and demographic characteristics.

```{r setup}
#| echo: false
#| eval: true
source(here::here("R", "utils.R"))
source(here::here("R", "enrollment.R"))
```


::: {.callout-note appearance="simple" collapse="true"}

## Original Medicare Enrollees

Medicare beneficiaries enrolled in the traditional health care system (or Fee-for-Service).

:::


::: {.callout-note appearance="simple" collapse="true"}

## Medicare Advantage and Other Health Plan Enrollees

Medicare beneficiaries enrolled in health plans that are offered by private companies approved by Medicare to provide health care coverage offered at a uniform premium and uniform level of cost-sharing to all Medicare beneficiaries residing in the service area (or segment of the service area). Each type of plan has special rules and exceptions.

:::

::: {.panel-tabset}

## Yearly Enrollment

```{r mdcr_enroll_ab_1}
enroll_1 |> 
  dplyr::select(Year = year, 
                Total = total, 
                Original = original, 
                Advantage = ma_other) |> 
  gt::gt(rowname_col = "Year") |> 
  gt::sub_missing() |> 
  gt::fmt_number(columns = c("Total", "Original", "Advantage"),
                 suffixing = TRUE,
                 decimals = 2) |>
  gt::opt_stylize(style = 1)
```

## Absolute Change

```{r mdcr_enroll_ab_1_2}
enroll_1 |> 
  dplyr::select(Year = year, 
                Total = total_chg, 
                Original = original_chg, 
                Advantage = ma_other_chg) |> 
  gt::gt(rowname_col = "Year") |> 
  gt::sub_missing() |> 
  gt::fmt_number(columns = c("Total", "Original", "Advantage"),
                 force_sign = TRUE,
                 decimals = 2,
                 suffixing = TRUE) |>
  gt::opt_stylize(style = 1)
```

## Relative Change

```{r mdcr_enroll_ab_1_3}
enroll_1 |> 
  dplyr::select(Year = year, 
                Total = total_pct_chg, 
                Original = original_pct_chg, 
                Advantage = ma_other_pct_chg) |> 
  gt::gt(rowname_col = "Year") |> 
  gt::sub_missing() |> 
  gt::fmt_percent(columns = c("Total", "Original", "Advantage"),
                  force_sign = TRUE) |> 
  gt::opt_stylize(style = 1)
```

:::

```{r}
#| label: fig-origmapct
#| fig-cap: "Breakdown of percentage of Medicare types from 2008 to 2021"
#| warning: false
#| echo: false
#| eval: false

enroll_1 |> 
  dplyr::select(Year = year, 
                #total, 
                Original = original, 
                "Medicare Advantage" = ma_other) |> 
  tidyr::pivot_longer(!Year, 
                      names_to = "Type", 
                      values_to = "Enrollees") |> 
  ggplot2::ggplot(ggplot2::aes(Year, 
                               Enrollees, 
                               group = Type, 
                               fill = Type)) +
  ggchicklet::geom_chicklet(width = 0.75) +
  ggplot2::scale_fill_manual(
    name = NULL,
    values = c(
      "Medicare Advantage" = "#ae4544",
      "Economy" = "#d8cb98",
      "Climate Change" = "#a4ad6f",
      "Original" = "#cc7c3a",
      "Healthcare" = "#436f82",
      "Foreign Policy" = "#7c5981",
      "Other" = "#cccccc"
    )
    #breaks = setdiff(unique(debates2019$topic), "Other")
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(nrow = 1)
  ) +
  ggplot2::coord_flip() +
  ggplot2::labs(
    x = NULL, 
    y = NULL, 
    fill = NULL
  ) +
  hrbrthemes::theme_ipsum_rc(grid="X") +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
    color = "gray60", 
    size = 10)) +
  ggplot2::theme(legend.position = "top")
```

```{r}
#| warning: false
#| echo: false
#| eval: false
enroll_1 |> 
  dplyr::select(year, 
                original_pct, 
                ma_other_pct) |> 
  tidyr::pivot_longer(!year, 
                      names_to = "measure", 
                      values_to = "data") |> 
  dplyr::group_by(year) |> 
  dplyr::summarise(list_data = list(data)) |> 
  gt::gt() |> 
  gtExtras::gt_plt_bar_stack(column = list_data,
                             labels = c("Original ", 
                                        " Advantage"),
                             palette = c("#ff4343", "#0a1c2b"),
                             fmt_fn = scales::label_percent(accuracy = 0.1)) |> 
  gtExtras::gt_theme_538()
```

```{r}
#| warning: false
#| echo: false
#| eval: false
enroll_1 |> 
  dplyr::select(year, 
                total_pct_chg,
                original_pct_chg, 
                ma_other_pct_chg) |> 
  provider:::gt_prov(divider = c("year", 
                                 "ma_other_pct_chg"),
                     pctchg = c("total_pct_chg", 
                                "original_pct_chg", 
                                "ma_other_pct_chg")) |>
  gt::tab_spanner(label = "YEAR-OVER-YEAR CHANGE (%)",
                  columns = c(total_pct_chg,
                              original_pct_chg, 
                              ma_other_pct_chg)) |> 
  gt::cols_label(total_pct_chg = "Total",
                 original_pct_chg = "Original", 
                 ma_other_pct_chg = "Advantage")
```

<br>

```{r}
enroll_1 |> 
  dplyr::select(year,
                total,
                original,
                ma_other,
                total_chg,
                original_chg, 
                ma_other_chg) |> 
  dplyr::summarise(n = dplyr::n(),
                   avg_total = mean(total),
                   avg_original = mean(original),
                   avg_ma_other = mean(ma_other),
                   tot_chg_sum = sum(total_chg, na.rm = TRUE),
                   avg_tot_chg = tot_chg_sum / n,
                   orig_chg_sum = sum(original_chg, na.rm = TRUE),
                   avg_orig_chg = orig_chg_sum / n,
                   ma_chg_sum = sum(ma_other_chg, na.rm = TRUE),
                   avg_ma_chg = ma_chg_sum / n) |> 
  dplyr::select(!c(n, dplyr::contains("sum"))) |> 
  pct(avg_original, avg_total) |> 
  pct(avg_ma_other, avg_total) |> 
  dplyr::glimpse()
```

<br>

## Enrollment by Area of Residence, 2013-2021

::: {.callout-note appearance="simple" collapse="true"}

## Metropolitan Statistical Area

A Core Based Statistical Area associated with at least one urbanized area that has a population of at least 50,000. The Metropolitan Statistical Area comprises the central county or counties containing the core, plus adjacent outlying counties having a high degree of social and economic integration with the central county or counties as measured through commuting.

:::


::: {.callout-note appearance="simple" collapse="true"}

## Micropolitan Statistical Area

A Core Statistical Area associated with at least one urban cluster that has a population of at least 10,000, but less than 50,000. The Micropolitan Statistical Area comprises the central county or counties containing the core, plus adjacent outlying counties having a high degree of social and economic integration with the central county or counties as measured through commuting.

:::


::: {.callout-note appearance="simple" collapse="true"}

## Non-Core-Based Statistical Area

All remaining counties, often labeled “noncore” counties because they are not part of “core-based” metropolitan or micropolitan areas.

:::

```{r mdcr_enroll_ab_2}
enroll_2
```

<br>

### Percentage of Population Enrolled in Medicare

```{r}
#| column: page
enroll_2 |> 
  dplyr::select(year,
                residence = area_of_residence,
                Total = total_pct) |> 
  tidyr::pivot_wider(names_from = "year", 
                     values_from = "Total") |> 
  gt::gt(rowname_col = "residence") |> 
  gt::tab_header(
    title = gt::md("**Medicare Enrollment by Area of Residence**"),
    subtitle = gt::md("**Percentage of Population Enrolled in Medicare, 2013 - 2021**")
  ) |> 
  gt::cols_align(align = "right", 
                 columns = residence) |> 
  gt::cols_align_decimal() |> 
  # gt::opt_table_font(
  #   font = c(
  #     gt::google_font(name = "Karla"),
  #     gt::default_fonts())) |>
  gt::sub_missing() |>
  gt::fmt_percent(columns = 2:10,
                  decimals = 1,
                  drop_trailing_zeros = TRUE,
                  drop_trailing_dec_mark = TRUE) |>
  gt::data_color(direction = "row",
                 method = "numeric",
                 palette = "Blues",
                 reverse = TRUE) |>
  gt::tab_source_note(
    source_note = gt::md(
    "**Source**: [CMS Program Statistics - Medicare Total Enrollment](https://data.cms.gov/summary-statistics-on-beneficiary-enrollment/medicare-and-medicaid-reports/cms-program-statistics-medicare-total-enrollment).")) |> 
  gt::opt_stylize(style = 1, add_row_striping = FALSE)
```

<br>

### Percentage Enrolled in Original Medicare or Medicare Advantage

```{r}
#| column: page
#| echo: false
#| eval: false
enroll_2 |> 
  dplyr::select(year,
                residence = area_of_residence,
                ORIG = original_pct,
                MA = ma_other_pct) |> 
  tidyr::pivot_longer(cols = ORIG:MA, 
                      names_to = "Type", 
                      values_to = "Percentage") |> 
  tidyr::unite("year_type", c(year, Type)) |> 
  tidyr::pivot_wider(names_from = "year_type", 
                     values_from = "Percentage") |> 
  gt::gt(rowname_col = "residence") |> 
  gt::tab_spanner_delim(delim = "_") |>
  gt::cols_align(align = "right", 
                 columns = residence) |> 
  gt::cols_align_decimal() |> 
  gt::opt_table_font(
    font = c(
      gt::google_font(name = "Karla"),
      gt::default_fonts())) |>
  gt::sub_missing() |>
  gt::fmt_percent(columns = 2:19,
                  decimals = 1,
                  drop_trailing_zeros = TRUE,
                  drop_trailing_dec_mark = TRUE) |>
  gt::data_color(direction = "row",
                 method = "numeric",
                 palette = "Blues") |>
  gt::opt_vertical_padding(scale = 1.65) |>
  gt::opt_stylize(style = 1, add_row_striping = FALSE)
```

```{r}
enroll_2 |> 
  dplyr::select(year,
                residence = area_of_residence,
                Original = original_pct,
                Advantage = ma_other_pct) |> 
  tidyr::pivot_longer(cols = Original:Advantage, 
                      names_to = "Type", 
                      values_to = "Percentage") |> 
  tidyr::pivot_wider(names_from = "year",
                     values_from = "Percentage") |>
  gt::gt(rowname_col = "Type", 
         groupname_col = "residence") |> 
  gt::opt_stylize(style = 1, 
                  add_row_striping = FALSE) |>
  gt::tab_header(
    title = gt::md("**Medicare Enrollment by Area of Residence**"),
    subtitle = gt::md("**Percentage of Medicare Beneficiaries Enrolled in *Original Medicare* or *Medicare Advantage*, 2013 - 2021**")
  ) |> 
   gt::tab_options(
     column_labels.vlines.color = "blue",
    column_labels.font.weight = "bold",
    stub.font.weight = "bold",
    stub.font.size = gt::pct(80),
    row_group.font.weight = "bold",
    row_group.font.size = gt::pct(70),
    row_group.as_column = TRUE) |>
  gt::cols_align(align = "right", 
                 columns = c("residence", "Type")) |> 
  gt::cols_align(align = "center", 
                 columns = 3:11) |> 
  gt::cols_align_decimal() |> 
  # gt::opt_table_font(
  #   font = c(
  #     gt::google_font(name = "Karla"),
  #     gt::default_fonts())) |>
  gt::sub_missing() |>
  gt::fmt_percent(columns = 3:11,
                  decimals = 1,
                  drop_trailing_zeros = TRUE,
                  drop_trailing_dec_mark = TRUE) |>
  gt::data_color(direction = "row",
                 method = "numeric",
                 palette = "Blues") |>
  # gt::tab_style(
  #   style = gt::cell_fill(color = "lightblue"),
  #   locations = gt::cells_row_groups()) |> 
  gt::opt_vertical_padding(scale = 0.65) |>
  gt::tab_source_note(
    source_note = gt::md(
    "**Source**: [CMS Program Statistics - Medicare Total Enrollment](https://data.cms.gov/summary-statistics-on-beneficiary-enrollment/medicare-and-medicaid-reports/cms-program-statistics-medicare-total-enrollment)."))
```

<br>


### Enrollment Averages

```{r mdcr_enroll_avgs}
enroll_avgs
```