---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse  = FALSE,
  echo      = TRUE,
  message   = FALSE, 
  warning   = FALSE,
  error     = TRUE,
  comment   = "#>",
  dpi       = 300, 
  out.width = "100%",
  fig.path  = "man/figures/README-"
)
options(scipen = 999)
#library(timetk)
#library(ggplot2)
#library(plotly)
#library(lubridate)
```

```{r rds_files, include = FALSE}
mdcr_enroll_ab_1 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_1.rds"))
mdcr_enroll_ab_2 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_2.rds"))
mdcr_enroll_ab_3 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_3.rds"))
mdcr_enroll_ab_4 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_4.rds"))
mdcr_enroll_ab_5 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_5.rds"))
mdcr_enroll_ab_6 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_6.rds"))
mdcr_enroll_ab_7 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_7.rds"))
mdcr_enroll_ab_8 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_8.rds"))

# mdcr_enroll_avgs <- readr::read_rds(here::here("data", "mdcr_enroll_avgs.rds"))

mdcr_enroll_ab_33 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_33.rds"))
mdcr_enroll_ab_34 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_34.rds"))
mdcr_enroll_ab_35 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_35.rds"))
mdcr_enroll_ab_36 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_36.rds"))
mdcr_enroll_ab_37 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_37.rds"))

# mdcr_enroll_ab_38 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_38.rds"))
# mdcr_enroll_ab_39 <- readr::read_rds(here::here("data", "mdcr_enroll_ab_39.rds"))
```

# CMS Program Statistics

The CMS Program Statistics (CPS) includes high-level summary reports focused on Medicare populations, use, and payment, as well as counts for Medicare-certified institutional and non-institutional providers.

## Medicare Enrollment

The Medicare Enrollment reports contain trend, demographic, and state tables showing total Medicare enrollment, Original Medicare enrollment, Medicare Advantage and Other Health Plan enrollment, newly-enrolled beneficiaries, deaths, Medicare-Medicaid Enrollment, and Medicare Part D enrollment.

### Medicare Total Enrollment

Information on all Medicare enrollees by geography, type of coverage, type of entitlement, and demographic characteristics.

```{r eval=FALSE}
mdcr_enroll_ab_1 |> 
  dplyr::select(year, 
                original_pct, 
                ma_other_pct) |> 
  tidyr::pivot_longer(!year, 
                      names_to = "measure", 
                      values_to = "data") |> 
  dplyr::group_by(year) |> 
  dplyr::summarise(list_data = list(data)) |> 
  gt::gt() |> 
  gtExtras::gt_plt_bar_stack(column = list_data,
                             labels = c("Original ", 
                                        " Advantage"),
                             palette = c("#ff4343", "#0a1c2b"),
                             fmt_fn = scales::label_percent(accuracy = 0.1)) |> 
  gtExtras::gt_theme_538()
```

<br>

```{r eval=FALSE}
mdcr_enroll_ab_1 |> 
  dplyr::select(year, 
                total_pct_chg,
                original_pct_chg, 
                ma_other_pct_chg) |> 
  provider:::gt_prov(divider = c("year", 
                                 "ma_other_pct_chg"),
                     pctchg = c("total_pct_chg", 
                                "original_pct_chg", 
                                "ma_other_pct_chg")) |>
  gt::tab_spanner(label = "YEAR-OVER-YEAR CHANGE (%)",
                  columns = c(total_pct_chg,
                              original_pct_chg, 
                              ma_other_pct_chg)) |> 
  gt::cols_label(total_pct_chg = "Total",
                 original_pct_chg = "Original", 
                 ma_other_pct_chg = "Advantage")
```

<br>

```{r eval=FALSE}
mdcr_enroll_ab_1 |> 
  dplyr::select(year,
                total,
                original,
                ma_other,
                total_chg,
                original_chg, 
                ma_other_chg) |> 
  dplyr::summarise(n = dplyr::n(),
                   avg_total = mean(total),
                   avg_original = mean(original),
                   avg_ma_other = mean(ma_other),
                   tot_chg_sum = sum(total_chg, na.rm = TRUE),
                   avg_tot_chg = tot_chg_sum / n,
                   orig_chg_sum = sum(original_chg, na.rm = TRUE),
                   avg_orig_chg = orig_chg_sum / n,
                   ma_chg_sum = sum(ma_other_chg, na.rm = TRUE),
                   avg_ma_chg = ma_chg_sum / n) |> 
  dplyr::select(!c(n, dplyr::contains("sum"))) |> 
  pct(avg_original, avg_total) |> 
  pct(avg_ma_other, avg_total)
```
<br>

Total Enrollment and Population by Area of Residence, 2013-2021

```{r mdcr_enroll_ab_2}
mdcr_enroll_ab_2 |> 
  dplyr::filter(area_of_residence == "United States") |> 
  dplyr::select(year, total_pct)
```

Part A and/or Part B Total, Aged, and Disabled Enrollees, 2008-2021

```{r mdcr_enroll_ab_3}
mdcr_enroll_ab_3
```

Part A and/or Part B Enrollees by Age Group, 2008-2021

```{r mdcr_enroll_ab_4}
mdcr_enroll_ab_4
```

Part A and/or Part B Enrollees by Demographics, 2013-2021

```{r mdcr_enroll_ab_5}
mdcr_enroll_ab_5
```

Part A and/or Part B Enrollees by Type of Entitlement and Demographics,
2013-2021

```{r mdcr_enroll_ab_6}
mdcr_enroll_ab_6
```

Part A and/or Part B Total, Aged, and Disabled Enrollees by Area of
Residence, 2013-2021

```{r mdcr_enroll_ab_7}
mdcr_enroll_ab_7
```

Part A and/or Part B Enrollees, by Type of Entitlement and Area of
Residence, 2013-2021

```{r mdcr_enroll_ab_8}
mdcr_enroll_ab_8
```

## Medicare Deaths

Information on all Medicare beneficiary deaths, by type of entitlement, health care delivery, demographic characteristics, and geography.

Total Beneficiaries by Month of Death, 2008-2021

```{r mdcr_enroll_ab_33}
mdcr_enroll_ab_33
```

Total Beneficiaries by Demographics, 2013-2021

```{r mdcr_enroll_ab_34}
mdcr_enroll_ab_34
```

Total Beneficiaries by Area of Residence, 2013-2021

```{r mdcr_enroll_ab_35}
mdcr_enroll_ab_35
```

Original Medicare Beneficiaries by Month of Death, 2008-2021

```{r mdcr_enroll_ab_36}
mdcr_enroll_ab_36
```

Original Medicare Beneficiaries by Area of Residence, 2013-2021

```{r mdcr_enroll_ab_37}
mdcr_enroll_ab_37
```

Medicare Advantage and Other Health Plan Beneficiaries by Month of
Death, 2008-2021

```{r mdcr_enroll_ab_38}
#mdcr_enroll_ab_38
```

Medicare Advantage and Other Health Plan Beneficiaries by Area of
Residence, 2013-2021

```{r mdcr_enroll_ab_39}
#mdcr_enroll_ab_39
```
